Automatic Identification and Annotation of Tasks in Structured Programs

### 2018-01-24 docs/ACM/paper.tex
worked on section 3.1 of the paper

\paragraph{The annotation zoo.}
Our equipment of choice to reveal task parallelism in programs is OpenMP 3.0.
This annotation system provides us with a set of pragmas that indicate to compiler
and runtime environment when and how tasks must be created.
We describe below which annotations we are using, and explain, informally, their
semantics.
Full overviews of the syntax and semantics of these annotations is publicly
available\footnote{For a quick overview, we refer the reader to the leaflet
``Summary of OpenMP 3.0 C/C++ Syntax", which is made available by the OpenMP
group.}; hence, we shall not dive into their details.
%
\begin{compactitem}
\item \textsf{parallel} (\textit{clauses}): forms a team of threads
which will execute the marked program region in parallel.
\item \textsf{single} (\textit{clauses}): specifies that a
program region must be executed by one thread in the team.
\item \textsf{task} (\textit{clauses}): creates a new task.
\item \textsf{taskwait}: defines a synchronization point where
threads must wait for the completion of child tasks.
\end{compactitem}
%
Some pragmas are associated with lists of clauses.
There are several of these clauses available in OpenMP 3.0; however, we only use
the following:
\begin{compactitem}
\item \textsf{default([shared/private])}: indicates that variables are either
shared or replicated among tasks.
\item \textsf{firstprivate($v$)}: indicates that $v$ must be replicated among
tasks, and initialized with its value at the point where the annotation is
executed.
\item \textsf{untied}: if a task has this modifier, then its associated code can
be executed by more than one thread; e.g., threads might alternate execution due
to preemption.
\item \textsf{depend}($\mathit{in}/\mathit{out}/\mathit{inout}$):
determine if data is read ($\mathit{in}$), written ($\mathit{out}$) or both
within a task region; thus, effectively setting dependences among tasks.
\item \textsf{depend}({\em condition}): this clause defines the conditions
under which a task is allowed to create threads.
We use it to enforce the cost model.
\end{compactitem}

@@

The algorithm to find tasks is illustrated in Figure \ref{fig:alg_main}. It uses
information of two abstract program structures: the \emph{Control Flow Graph (CFG)} and
the \emph{Program Dependence Graph (PDG)}. The key intuition behind
finding potential tasks relies on a concept that we call \emph{Helices}. A \emph{Helix}
is a set of nodes in the PDG that depend on a Strong Connected Component (SCC) but \emph{do not
integrate the SCC}. Strong Connected Components in the Dependence Graph represent statically
detected iteration dependence, so it arises with loops or with recursive patterns (when the Graph is
interprocedural).  Figure~\ref{fig:helix}  shows what represents a \emph{Helix}.
The insight behind \emph{Helices} is that they represent instructions that are deemed
to repeat more than once and, since they do not have a cycle, they can be executed in parallel.

Therefore, the first step in the algorithm to find tasks is to map these \emph{Helices}
into program regions. The algorithm goes through the Dependence Graph looking for \emph{Helices}
and then finds the minimal covering region for each \emph{Helix}. These minimal regions
are later going to be annotated as tasks in the original program, should they fit a cost model.

After finding potential task regions, next is the \emph{expansion} step. We expand the region to a larger 
one as long as two requisites are met:

\begin{enumerate}  
\item Input and output dependencies are never increased;
\item The potential parallelism is never reduced.
\end{enumerate}

@@

\begin{itemize}
\item Performance
\item Optimizations
\item Versatility
\item Applicability
\item Practicality
\end{itemize}

@@

%APPLICABILITY
%We'll show TaskMiner application on larger benchmarks, such as those of spec or those of the LLVM's test-suite.
\textbf{\textit{Versatility}}.

>>>>

\paragraph{The annotation zoo.}
Our equipment of choice to reveal task parallelism in programs is OpenMP 4.0.
This annotation system provides us with a set of pragmas that indicate to compiler
and runtime environment when and how tasks must be created.
We describe below which annotations we are using, and explain, informally, their
semantics.
Full overviews of the syntax and semantics of these annotations is publicly
available\footnote{For a quick overview, we refer the reader to the leaflet
``Summary of OpenMP 4.0 C/C++ Syntax", which is made available by the OpenMP
group.}; hence, we shall not dive into their details.
%
\begin{compactitem}
\item \textsf{parallel} (\textit{clauses}): forms a team of threads
which will execute the marked program region in parallel.
\item \textsf{single} (\textit{clauses}): specifies that a
program region must be executed by one thread in the team.
\item \textsf{task} (\textit{clauses}): creates a new task.
\item \textsf{taskwait}: defines a synchronization point where
threads must wait for the completion of child tasks.
\end{compactitem}
%
Some pragmas are associated with lists of clauses.
There are several of these clauses available in OpenMP 4.0; however, we only use
the following:
\begin{compactitem}
\item \textsf{default([shared/private])}: indicates that variables are either
shared or replicated among tasks.
\item \textsf{firstprivate($v$)}: indicates that $v$ must be replicated among
tasks, and initialized with its value at the point where the annotation is
executed.
\item \textsf{untied}: if a task has this modifier, then its associated code can
be executed by more than one thread; e.g., threads might alternate execution due
to preemption and load balancing.
\item \textsf{depend}($\mathit{in}/\mathit{out}/\mathit{inout}$):
determine if data is read ($\mathit{in}$), written ($\mathit{out}$) or both
within a task region; thus, effectively setting dependences among tasks.
\item \textsf{if}({\em condition}): this clause defines the conditions
under which a task is allowed to create threads.
We use it to enforce the cost model.
\end{compactitem}

@@

The algorithm to find tasks is illustrated in Figure \ref{fig:alg_main}. It uses
information of two abstract program structures: the \emph{Control Flow Graph (CFG)} and
the \emph{Program Dependence Graph (PDG)}. The key intuition behind
finding potential tasks relies on a concept that we call \emph{Helices}.

\begin{definition}[Helix]
\label{def:helix}
Given a set $S$ of strongly connected nodes in the Dependence Graph $G$, a set of nodes $H$ is a Helix if, and only if:
\begin{itemize}
\item $H \subset \bigcup_i N_i[S]$;
\item $H  \cap S = \emptyset$;
\item $H$ is connected and maximal.
\end{itemize}
\end{definition}

We can view a \emph{Helix} as a set of nodes in the PDG that depends on a Strongly Connected Component
(SCC) but \emph{do not integrate the SCC}. That means helices are all those subgraphs that depend on
the SCC and do not form a cyclic dependence with the SCC itself.

Statically, a Strongly Connected Component in the PDG represents iteration dependence, so it
arises due to loops or recursive patterns, should the PDG is interprocedural. The insight behind 
\emph{Helices} is that they represent instructions that are deemed to be executed more than once
for they depend on a cyclic dependence (the SCC), and, since they do not integrate the SCC, they can
be executed in parallel.

Thus, the first step in the algorithm to find tasks is to map these \emph{Helices}
into program regions. The algorithm goes through the Dependence Graph looking for \emph{Helices}
and then finds the minimal covering region for each \emph{Helix}. These minimal regions
are later going to be annotated as tasks in the original program, should they fit a cost model.

After finding potential task regions, the next critical step is the \emph{expansion}.

\begin{definition}[Region expansion]
\label{def:expansion}
A task region $R$ can be expanded into a larger region $R'$ if, and only if:
\begin{itemize}
\item $|M(R')| \leq |M(R)|$, in which $M(R)$ and $M(R')$ are the set of input and output memory
dependences in regions $R$ and $R'$ respectively.
\item $R'$ is still inside a loop or a cyclic construct.
\end{itemize}
\end{definition}

@@

\begin{compactitem}
\item \textsf{Performance}: we run \Taskminer on the BOTS Benchmark and show that the programs annotated by taskminer are sometimes as fast as programs annotated manually, and never slower than their sequential counterparts, thus proving it to be a conservative and useful technique. 
\item \textsf{Optimizations}: we show some optimizations that can be applied to \Taskminer in order to improve performance of the annotated programs.
\item \textsf{Versatility}: we demonstrate \Taskminer's versatility when it is able to find tasks in several different program paradigms.
\item \textsf{Applicability}: we run \Taskminer on larger benchmarks to show the potential of finding task parallelism in big applications.
\item \textsf{Practicality}: here we present \Taskminer's runtime and show that it can be easily used by a programmer to find task parallelism in applications. We evidence \Taskminer's practicality when compared to the method of manually finding parallelism, determining data dependences annotating the program.
\end{compactitem}

@@

%APPLICABILITY
%We'll show TaskMiner application on larger benchmarks, such as those of spec or those of the LLVM's test-suite.
\textbf{\textit{Applicability}}.
